---
title: Creating Publication-Quality Graphics with ggplot2
teaching: 2
exercises: 3
source: Rmd
---

::::::::::::::::::::::::::::::::::::::: objectives

- Learn how to use the {ggplot2} package to generate publication-quality graphics.
- Manipulate the aesthetics of a plot using different colors, shapes, and lines.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I create publication-quality graphics in R?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
library(gapminder)
```

Plotting our data is one of the best ways to
quickly explore it and the various relationships
between variables.

Today we'll be learning about the ggplot2 package, built on the `grammar of graphics`:

- **data set**,
- **mapping aesthetics**, 
- graphical **layers**



Let's start off building an example using the `gapminder` data from earlier.


```{r blank-ggplot, message=FALSE, fig.alt="Blank plot, before adding any mapping aesthetics to ggplot()."}
library("ggplot2")
ggplot(data = gapminder)
```


Now we're going to add in the **mapping aesthetics** using the
`aes` function. `aes` tells `ggplot` how variables in the **data**
map to *aesthetic* properties of the figure, such as which columns
of the data should be used for the **x** and **y** locations.

```{r ggplot-with-aes, message=FALSE, fig.alt="Plotting area with axes for a scatter plot of life expectancy vs GDP, with no data points visible."}
ggplot(data = gapminder, 
       mapping = aes(x = gdpPercap, y = lifeExp))
```

The final part of making our plot is to tell `ggplot` how we want to
visually represent the data. We do this by adding a new **layer**
to the plot using one of the **geom** functions.

```{r lifeExp-vs-gdpPercap-scatter, message=FALSE, fig.alt="Scatter plot of life expectancy vs GDP per capita, now showing the data points."}
ggplot(data = gapminder, 
       mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```


Modify the example so that the figure shows how life expectancy has
changed over time:

```{r, eval=T}
ggplot(data = gapminder, 
       mapping = aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point()
```


## Layers

Using a scatterplot probably isn't the best for visualizing change over time.
Instead, let's tell `ggplot` to visualize the data as a line plot:

```{r lifeExp-line}
ggplot(data = gapminder, 
       mapping = aes(x=year, y=lifeExp, group=country,color=continent)) +
  geom_line()+ 
  geom_point()
```



In this example, the *aesthetic* mapping of **color** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies
to the points. Now we can clearly see that the points are drawn on top of the
lines.

:::::::::::::::::::::::::::::::::::::::::  callout

## Tip: Setting an aesthetic to a value instead of a mapping

So far, we've seen how to use an aesthetic (such as **color**) as a *mapping* to a variable in the data. For example, when we use `geom_line(mapping = aes(color=continent))`, ggplot will give a different color to each continent. But what if we want to change the color of all lines to blue? You may think that `geom_line(mapping = aes(color="blue"))` should work, but it doesn't. Since we don't want to create a mapping to a specific variable, we can move the color specification outside of the `aes()` function, like this: `geom_line(color="blue")`.


::::::::::::::::::::::::::::::::::::::::::::::::::





## Transformations and statistics

ggplot2 also makes it easy to overlay statistical models over the data. To
demonstrate we'll go back to our first example:


Currently it's hard to see the relationship between the points due to some strong
outliers in GDP per capita. We can change the scale of units on the x axis using
the *scale* functions. These control the mapping between the data values and
visual values of an aesthetic. We can also modify the transparency of the
points, using the *alpha* function, which is especially helpful when you have
a large amount of data which is very clustered.

```{r axis-scale, fig.cap="Scatterplot of GDP vs life expectancy showing logarithmic x-axis data spread"}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) + scale_x_log10()
```

The `scale_x_log10` function applied a transformation to the coordinate system of the plot, so that each multiple of 10 is evenly spaced from left to right. For example, a GDP per capita of 1,000 is the same horizontal distance away from a value of 10,000 as the 10,000 value is from 100,000. This helps to visualize the spread of the data along the x-axis.

:::::::::::::::::::::::::::::::::::::::::  callout

## Tip Reminder: Setting an aesthetic to a value instead of a mapping

Notice that we used `geom_point(alpha = 0.5)`. As the previous tip mentioned, using a setting outside of the `aes()` function will cause this value to be used for all points, which is what we want in this case. But just like any other aesthetic setting, *alpha* can also be mapped to a variable in the data. For example, we can give a different transparency to each continent with `geom_point(mapping = aes(alpha = continent))`.


::::::::::::::::::::::::::::::::::::::::::::::::::





[base]: https://www.statmethods.net/graphs/index.html
[lattice]: https://www.statmethods.net/advgraphs/trellis.html
[ggplot2]: https://www.statmethods.net/advgraphs/ggplot2.html
[cheat]: https://www.rstudio.org/links/data_visualization_cheat_sheet
[cheat_all]: https://www.rstudio.com/resources/cheatsheets/
[ggplot-doc]: https://ggplot2.tidyverse.org/reference/


:::::::::::::::::::::::::::::::::::::::: keypoints

- Use `ggplot2` to create plots.
- Think about graphics in layers: aesthetics, geometry, statistics, scale transformation, and grouping.

::::::::::::::::::::::::::::::::::::::::::::::::::
